---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
library(ggplot2)
```

3.  Make a repository in the <https://github.com/stat545ubc-2023> Organization. You can do this by following the steps found on canvas in the entry called [MDA: Create a repository](https://canvas.ubc.ca/courses/126199/pages/mda-create-a-repository). One completed, your repository should automatically be listed as part of the stat545ubc-2023 Organization.

# Instructions

## For Both Milestones

-   Each milestone has explicit tasks. Tasks that are more challenging will often be allocated more points.

-  Each milestone will be also graded for reproducibility, cleanliness, and coherence of the overall Github submission. 

- While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

- The milestones will be equally weighted. 

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 36 points: 30 for your analysis, and 6 for overall reproducibility, cleanliness, and coherence of the Github submission. 

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset 

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 **(1 point)** Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: CHOICE_1_vancouver_trees
2: CHOICE_2_cancer_sample
3: CHOICE_3_apt_buildings
4: CHOICE_4_parking_meters

<!----------------------------------------------------------------------------->

1.2 **(6 points)** One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

```{r}
vancouver_trees %>%
  glimpse() %>%
  class()
```
```{r}
cancer_sample %>%
  glimpse() %>%
  class()
```
```{r}
apt_buildings %>%
  glimpse() %>%
  class()
```
```{r}
parking_meters %>%
  glimpse() %>%
  class()
```

<!----------------------------------------------------------------------------->

1.3 **(1 point)** Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 1. What lead you to choose this one? Briefly explain your choice below.

<!-------------------------- Start your work below ---------------------------->
## First Choice Dataset

I chose apt_buildings as my main choice because it has many columns that contain both numerical and non-numerical data. These can be used to apply various dplyr functions to analyze the data (such as group_by, filter, select, and summarize). I can also plot this data easily on ggplot functions. I am also really interested in this, as I am hoping to buy an apartment soon, and want to analyze the data for my personal interest.

<!----------------------------------------------------------------------------->

1.4 **(2 points)** Time for a  final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with your dataset. Note it down below.

<!-------------------------- Start your work below ---------------------------->
## End Goal

I would like to explore the relationship between the date of the apartment buildings being built compared to their security and safety features. This includes the presence or absence of fire alarms, sprinkler systems, emergency power, and an exterior fire exit. 

<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset 

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 **(12 points)** Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. 

Make sure that you're using dplyr and ggplot2 rather than base R for this task. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

1.  Plot the distribution of a numeric variable. (done)
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot. (done)
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year... (done)
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!(done)
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).


2.2 **(4 points)** For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->

# Report Introduction

Hello, my name is Bradley Hoskin, and I am exploring the dataset *apt_buildings* using tidyverse packages. I will mainly be using dplyr and ggplot2 functions to analyze and plot the data for my interests. My main goal today is to explore the dataset, and find if there is any relation between the presence or absence of safety features in the apartment buildings, and the year the apartment buildings were built. This report will contain both explanations of my code, as well as explanations of the figures and visuals of the data within this report.

# Dataset and Libraries

The necessary packages needed for this report are below. I highly recommend loading these packages, or installing and then loading them if you do not have these packages already installed. This will allow you to run my code and reproduce the results in this report.
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(datateachr)
```

# Introduction of the Dataset

I first want to get a quick look at the *apt_buildings* dataset. I used the *glimpse* function in order to get a a quick look at the dataset. The glimpse function tells me how many rows and columns there are in the tibble dataset, as well as the names of all the columns and a few rows of data for the columns. This gives us information on the type of data (such as numerical or non-numerical) collected in the dataset. I also used the *class* function to find the attributes of the dataset, which shows us that the dataset is a tibble, which allows us to apply the dplyr functions that will produce a new tibble as an output.
```{r}
apt_buildings %>%
  glimpse() %>%
  class()
```
# Exploration of the Dataset

After using both the *glimpse* and *class* functions to get a quick look at the dataset, I wanted to dive a bit deeper and explore it. This report analyzes the *apt_buildings* dataset in 4 different ways.

### Apartment Building Size

Important factors that are often overlooked in regards to apartment building safety are the number of stories and number of units within a given building. These two variables are essential in the case of an emergency, such as a fire or earthquake. Apartment buildings with more units and/or more stories may have a higher risk of injury and fatality during an emergency. Below is a scatter plot that shows the number of units compared to the number of storeys for the apartment buildings from the dataset. The function *ggplot* creates the plot based off the information I give it, such as the variables for the x and y axes. The x-axis variable is *no_of_storeys* and the y-axis variable is *no_of_units*, with both variables containing numerical values. Each dot in the scatter plot below represents an apartment building from the *apt_building* dataset. I labelled this plot as *units_vs_storeys_plot* and used the *print* function to display the actual scatter plot.

From the scatter plot, we can see that almost all of the apartment buildings from the dataset have under 5,000 units, no matter how many stories the apartment building contains. We can also see a trend in the dataset where as the number of stories increases in a given apartment building, the number of units it has usually increases. This is not surprising, as an apartment building with more floors is very likely to have more units. However, this is not true for all apartment buildings, and varies within the dataset. 
```{r}
units_vs_storeys_plot <- ggplot(apt_buildings, aes(x = no_of_units, y = no_of_storeys)) +
  geom_point(aes(colour = 'red'),
             alpha = 0.5,
             size = 1)
print(units_vs_storeys_plot)

```

I also wanted to plot the distribution of the number of stories for the apartment buildings within the *apt_building* dataset. Below is a histogram that was also plotted by the *ggplot* function. The x-axis shows the number of stories for the apartments, and the y-axis shows the number of apartment buildings that have the specified number of stories shown on the x-axis. I labelled this histogram as *no_of_storeys_plot* and used the *print* function to display the histogram.

From the histogram below, we can see the vast majority of apartment buildings from the dataset have 10 stories or less, with about half of the apartment buildings having less than 5 stories.
```{r}
no_of_storeys_plot <- ggplot(apt_buildings, aes(x = no_of_storeys)) +
  geom_histogram()
print(no_of_storeys_plot)
```

### Safety Features

The next thing I wanted to explore was if there was a relationship between the year the apartment buildings were built, and the presence or absence of safety features in the apartments themselves. I first used the *select* function to select only the columns I was interested in. These columns included the *year_built*, *fire_alarm*, *sprinkler_system*, *emergency_power* and *no_of_storeys* columns of the *apt_buildings* dataset. I did this three times, and for each I used the *filter* function to filter the data entries/rows to show apartment buildings from the dataset that do not have certain safety features. The first one I filtered for apartment buildings with no fire alarm, the second one I filtered for apartment buildings with no sprinkler system, and the third I filtered for apartment buildings with no emergency power. Each of these codes produced a new tibble as an output, that shows the specified columns of interest, while simultaniously filtering apartment buildings with my required specifications. 

Looking at the first tibble which filtered the dataset for apartment buildings with no fire alarms, we can see that very few apartment buildings exist with no fire alarms. From the entire dataset, only 12 apartment buildings had no fire alarm, which was very good. However, things get a little more bleak as we explore the other tibbles. 

The second tibble was filtered for apartment buildings with no sprinkler systems. Approximately one third of the dataset does not have sprinkler systems.

The third tibble was filtered for apartment buildings with no emergency power. Over half of the apartment buildings from the dataset does not have emergency power.

These results are very alarming, and can cause a significant risk to people living in the apartment buildings in the case of an emergency, such as a fire or earthquake.
```{r}
apt_buildings %>%
  select(year_built, fire_alarm, sprinkler_system, emergency_power, no_of_storeys) %>%
  filter(fire_alarm == "NO")
apt_buildings %>%
  select(year_built, fire_alarm, sprinkler_system, emergency_power, no_of_storeys) %>%
  filter(sprinkler_system == "NO")
apt_buildings %>%
  select(year_built, fire_alarm, sprinkler_system, emergency_power, no_of_storeys) %>%
  filter(emergency_power == "NO")
```
### Accessible Units

The final thing I explored was the number of accessible units present in the apartment buildings from the dataset. This is an important feature of apartment buildings for people with disabilities, and are especially important during emergencies such as a fire, earthquake, or medical emergency. I used the *ggplot* function to plot the the number of barrier free accessible units in the apartment buildings from the dataset. I used the *geom_bar* to create a boxplot of the data. I named this boxplot *no_barrier_free_accessible_units_plot* and used the *print* function to display the boxplot. The results from the boxplot below show that the majority of apartment buildings from the dataset do not have any barrier free accessible units.
```{r}
no_barrier_free_accessible_units_plot <- ggplot(apt_buildings, aes(x = no_barrier_free_accessible_units)) +
  geom_bar(width = 0.8)
print(no_barrier_free_accessible_units_plot)
```

# Conclusion

Thank you very much for reading and reviewing this analytical report on the *apt_building* dataset. I still have much more to analyze in my next report on the dataset, which should be ready in a couple of weeks.

<!----------------------------------------------------------------------------->

# Task 3: Choose research questions 

**(4 points)** So far, you have chosen a dataset and gotten familiar with it through exploring the data. You have also brainstormed one research question that interested you (Task 1.4).  Now it's time to pick 4 research questions that you would like to explore in Milestone 2! Write the 4 questions and any additional comments below. 

<!--- *****START HERE***** --->
## Future Questions for Milestone 2

1. How accessible are these apartment buildings for people with disabilities? 
2. Does the overall accessibility of these apartment buildings change in comparison to when they were built?
3. Is there a relation between the kind of amenities (if any) and the year the apartment building was built?
4. Does the property type influence the number of safety features within the apartment building / is there a relation between the property type and the number of safety features? 

<!----------------------------->

# Overall reproducibility/Cleanliness/Coherence Checklist 

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. An example of a major continuity error is having a data set listed for Task 3 that is not part of one of the data sets listed in Task 1. 

## Error-free code (3 points) 

For full marks, all code in the document should run without error. 1 point deduction if most code runs without error, and 2 points deduction if more than 50% of the code throws an error. 

## Main README (1 point)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub.

Minimum contents of the README file:

-   In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
-   In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?

Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## Output (1 point)

All output is readable, recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

(0.5 point deduction if any of the above criteria are not met. 1 point deduction if most or all of the above criteria are not met.)

Our recommendation: right before submission, delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. Then, after your final commit and push to Github, CHECK on Github to make sure that everything looks the way you intended!

## Tagged release (0.5 points)

You've tagged a release for Milestone 1. 

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.